Weak interspecific interactions in a sagebrush steppe: Evidence from observational data, population models, and a removal experiment

### 2016-06-24 removals/manuscript/removal_ms.tex
revising ms.

@@

\begin{enumerate}
\item A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions, implying very little competitive release. 
\item We conducted a removal experiment to test the model's ability to predict competitive release. We established new quadrats in the same location where the historical data were collected. We assigned half the new quadrats to a  perennial grass removal treatment and half to a treatment in which we removed the dominant shrub, \textit{Artemisia tripartita}.
\item We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment. For survival and recruitment, the removal treatment effects were never significantly positive, indicating that competitive release was not underestimated by the interspecific effects included in the ``baseline'' model. 
\item For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive release. However, including information about the location of these grass individuals with respect to removed \textit{Artemisia} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
\item For three of the four species, individual based models and integral projections models  based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal. For \textit{P. spicata}, the population models that included effects of \textit{Artemisia} removal projected much stronger competitive release than the baseline models. 
\item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, overall the results increase our confidence that competitive release is in fact small for at least three of our four study species. The implication is that resource competition is not the primary factor determining the dynamics of this community.

@@

Stable coexistence requires that intraspecific interactions must limit populations more than interspecific interactions (Chesson 2000). In other words, when species are limited by different factors, intraspecific limitations will be stronger than interspecific limitations. For example, species might differ in their use of multiple limiting resources (Tilman), respond differently to spatial and temporal variation in the environment (Chesson 2000), or host different natural enemies (Janzen, Connell). Whenever we are studying populations of species that appear to be coexisting stably, such as populations that have co-occurred in close proximity for extended periods, we should expect intraspecific limitations to be stronger than interspecific limitations. However, we still might expect interspecific interactions to be strong and negative.

@@

We conducted a removal experiment in a sagebrush steppe to test whether our models, which are based on observational data, underestimate the strength of interspecific competition and competitive release. We studied the responses of three perennial grass species to removal of sagebrush and the response of sagebrush to removal of all grasses. We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs. removal treatments after accounting for differences in local species composition? Second, how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data?

@@

The U.S. Sheep Experiment Station (USSES) is located 9.6 km north of Dubois, Idaho (44.2°N, 112.1°W), 1500 m above sea level. During the period of data collection (1926 – 1957), mean annual precipitation was 270 mm and mean temperatures ranged from -8\degree C (January) to 21\degree C (July). The vegetation is dominated by the shrub, \textit{Artemisia tripartita}, and the C3  perennial bunchgrasses \textit{Pseudoroegneria spicata}, \textit{Hesperostipa comata}, and \textit{Poa secunda}. These four species, the focus of our models, accounted for over 70\% of basal cover (grasses) and 60\% of canopy cover (shrubs and forbs).

Scientists at the USSES established 26 1-m$^2$ quadrats between 1926 and 1932. Eighteen quadrats were distributed among four ungrazed exclosures, and eight were distributed in two paddocks grazed at medium intensity spring through fall. All quadrats were located on similar topography and soils. In most years until 1957, all individual plants in each quadrat were mapped using a pantograph (Blaisdell 1958). The historical data set is public and available online \citep*{zachmann-etal-2010}. In 2007, we relocated 14 of the original quadrats, all of which are located inside permanent livestock exclosures, and resumed annual mapped censusing using the traditional pantograph method.

We extracted data on survival, growth, and recruitment from the mapped quadrats based on plants' spatial locations. Our approach tracks genets, which may be composed of multiple polygons, as they fragment and/or coalesce. Each mapped polygon is classified as a surviving genet or a new recruit based on its spatial location relative to genets present in previous years (Lauenroth and Adler 2008). We modeled  vital rates using data from 22 year-to-year transitions between 1926 and 1957 and eight year-to-year transitions from 2007 to 2015. For the first two transitions in the 1920's only four quadrats were observed, while at least 14 quadrats were observed for all subsequent transitions.

\subsection*{Removal experiment}

In spring 2011, in a large exclosure containing six of the permanent, historical quadrats, we selected locations for an additional 16 quadrats for the removal treatments. We used the following criteria to select the quadrat locations. Each plot had to have xx? cumulative cover/density? of the three target perennial grass species as well as yy? cover/density of \textit{Artemisia}. We mapped the quadrats in June, 2011 and then implemented the removal treatments in September (?), 2011. We randomly assigned 8 of the sixteen new quadrats to the sagebrush removal treatment and the remaining 8 to the grass removal treatment. We cut sagebrush stems at ground level and applied herbicide to the cut stems. In the grass removal plots, we herbicided all perennial grasses. We removed plants zz cm beyond the quadrat boundaries to minimize edge effects. The sagebrush removal was virtually 100\% effective; grass removals required additional herbicide applications in April and May of each year,  with fewer grasses remaining each additional year.

@@

Although the primary focus of our analysis is comparison of observed removal responses with response predicted by our population models, we also conducted a more traditional statistical analysis of differences in cover trends on control vs. removal plots. The response variable for this analysis was year-to-year change in cover at the quadrat scale, calculated as $log(Cover_{t+1}/Cover_t)$. We tested for the effect of removal treatment on this metric of cover change with a linear mixed effects model implemented in the \textit{lme4} package (Bates et al. 2015) of \textit{R}, including year and quadrat as random effects. We considered the removal treatment effect significant when the 95\% confidence interval on the estimated coefficient did not overlap zero.

@@

Our regression models for survival, growth and recruitment follow previous work (Adler et al. 2010, Chu and Adler 2015) with a few exceptions, such as a new intercept term for the removal treatment.  We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (‘group’ here means a set of nearby quadrats located within one pasture or grazing exclosure), along with the removal treatment effect. More formally, the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatment $h$, from time $t$ to $t+1$  as
\begin{equation}
\mbox{logit}(S_{ijgh,t}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  \boldsymbol{\omega}_{j,t}^S \boldsymbol{W}_{ij,t}
\label{eqn:survReg}
\end{equation}
where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial group,$\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival, and $\boldsymbol{\omega}$ is a vector of per capita interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$ , by each species on the focal species. The vector $\boldsymbol{W}$ includes crowding from the four dominant species,  \textit{A. tripartita}, \textit{P. spicata}, \textit{H. comata}, and \textit{Poa secunda}, as well as two aggregated covariates representing crowding from (1)  all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points.

Our growth model has a similar structure. The change in genet size from time $t$ to $t+1$ , conditional on survival, is given by:
\begin{equation}
\mbox{logit}(u_{ijgh,t+1}) = \gamma_{j,t}^G + \varphi_{jg}^G+  \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} +  \boldsymbol{\omega}_{j,t}^G \boldsymbol{W}_{ij,t} + \varepsilon_{ij,t}^G .
\label{eqn:growReg}
\end{equation}
To capture non-constant error variance in growth, we modeled the variance $\varepsilon$  about the growth curve (\ref{eqn:growReg})  as a nonlinear function of predicted genet size:
\begin{equation}
Var(\varepsilon_{ij,t}^G) = a \exp ^{(bu_{ij,t+1})} .
\label{eqn:growVar}
\end{equation}

We model the crowding experienced by a focal genet as a function of the distance to and size of neighbor genets. In previous work, we assumed that the decay of crowding with neighbor distance followed a Gaussian function (Chu and Adler 2015), but here we use a data-driven approach (Teller et al. 2016). We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor areas across a set of concentric annuli, $k$, centered at the plant,
\begin{equation}
w_{ijm,k} = F_{jm}(d_{k})A_{i,k}      % can I put subscripts on a function?
\label{eqn:wik}
\end{equation}
where $F_{jm}$ is the competition kernel, $d_{k}$ is the average of the inner and outer radii of annulus $k$,
and $A_{im,k}$ is the total area of genets of species $m$
in annulus $k$ around genet $i$. The total crowding on genet $i$ exerted by species $m$ is
\begin{equation}
W_{ijm}  =\sum_k {w_{ijm,k}} .
\end{equation} 
Note that $W_{ijj}$ gives intraspecific crowding. The $W$'s are then introduced as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.

We assume that the competition kernel, $F(d)$, is non-negative and decreasing, which means that distant plants have less effect than close plants. Otherwise, we let the data dictate the shape of $F$ by fitting splines.  The demographic models such as \eqref{eqn:survReg} have $\gamma$, $\varphi$, $\chi$ , $\beta$, $\boldsymbol{\omega}$, and a vector of spline coefficients $\vec{b}$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R},  by making $\vec{b}$ the argument of an objective function that computes $W$ using the input value of $\vec{b}$, calls the model-fitting function
\texttt{lmer} and/or \texttt{glmer} to fit the other model parameters, and returns the negative log likelihood. We fit models for both survival and growth, and used the sum of the two log likelihoods. This assumes that one measure of crowding affects survival and growth. We estimated the $F$ functions using survival and growth models that depended only on intraspecific crowding. Based on previous work showing that competition kernels based on the effects of neighbors outperformed models based on the response of target plants (Adler et al. 2010), we set all $F_{jm}$ equal to $F_{jj}$. We used data from all historical and contemporary control treatment plots to estimate the competition kernels, which are shown in Fig. \ref{fig:CompKernels}.

Once we had estimated the parameters of the competitions kernels, we calculated the observed values of $W$ and fit the full survival and growth regressions, which include interspecific interaction coefficients, $\boldsymbol{\omega}$. All genets in a quadrat were included in calculating $W$, but plants located within 5 cm of quadrat edges were not used in fitting. We fit the models using the \textit{R-INLA} package for \texttt{R} (Rue et al. 20xx).

We model recruitment at the quadrat level rather than at the individual genet level because the mapped data do not allow us to determine which recruits were produced by which potential parent genets. We assume that the number of individuals, $y$, of species $j$ recruiting at time $t+1$ in the location $q$ follows a negative binomial distribution:
\begin{equation}
y_{jq,t+1}= NegBin(\lambda_{jq,t+1},\theta) 	   
\label{eqn:recrDataModel}
\end{equation}
where $\lambda$ is the mean intensity and $\theta$ is the size parameter. In turn,  $\lambda$ depends on the composition of the quadrat in the previous year:
\begin{equation}
\lambda_{jq,t+1} = C'_{jq,t} \exp{ (\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R +\boldsymbol{ \omega}^R \boldsymbol{\sqrt{C'}_{q,t}} ) }
\label{eqn:recrProcessModel}
\end{equation}
where $R$ refers to Recruitment, $C'_{jq,t}$ is the ‘effective cover’ (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ is a time-dependent intercept, $\varphi$ is a coefficient for the effect of group location, $\boldsymbol{\omega}$ is a vector of coefficients that determine the strength of intra- and interspecific density-dependence, and $\boldsymbol{C'}$ is the vector of ‘effective’ cover of each species in the community. Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth (Ricker 1954), except that we use the square root of cover in the density-dependent term. We chose not to directly use a Ricker equation as Adler et al. (2010) did, as recruitment models with the square-root transformation had smaller Deviance Information Criterion (DIC) values than the ones without the transformation (Spiegelhalter et al. 2002). Following previous work (Adler et al. 2010), we treated year and group variables as random factors, allowing intercepts to vary among years and spatial locations.

Due to the possibility that plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the group, $g$, in which the quadrat is located:
$C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$								
where $p$ is a mixing fraction between 0 and 1.

We ran Markov Chain Monte Carlo (MCMC) simulations implemented in WinBUGS 1.4 (Lunn et al. 2000) to estimate the recruitment model parameters. Each model was run for 20,000 iterations and 2 chains with different initial values for parameters. We discarded the initial 10,000 samples. Convergence was observed graphically for all parameters and diagnosed using the Brooks-Gelman statistic (Brooks and Gelman 1998).

For all three vital rate models, we considered the effect of removal treatments to be significant when the confidence interval on the estimate of $\chi$ did not overlap zero. In an effort to further understand the removal treatment effects, we explored additional variations on the growth model with a linear random effects model. For hypothesis testing with these models, we focused on the 95\% confidence interval of the coefficient of interest. We also compared the species composition of the control and removal plots at both the quadrat and neighborhood ($\boldsymbol{W}$) scales.

@@

The vital rate regressions provide a qualitative test of removal treatment effects: Does removal have an effect on a vital rate that goes above and beyond the effects of species interactions that are explicit in the model? However, the population-level response of a species requires integrating across all three vital rates. We used an individual-based model (IBM) to compare observed and predicted changes in population size (cover) from one year to the next.

To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the actual configuration of the four focal species (genet sizes and locations) observed in year $t$. We then projected the model forward one year, applying the random year effects corresponding to year $t$ and the spatial random effects corresponding to the quadrat of interest. We first applied the recruitment regression to determine the expected number of new recruits at time $t+1$, which we then multiplied by the mean size of recruits observed for the focal species to get expected cover of new recruits. We then used the survival regression to predict the survival probability of each genet, and the growth regression to predict the change in size of each genet. Expected cover at $t+1$ is the sum of the survival probabilities multiplied by the expected sizes plus the area of new recruits. 
% need to REDO with observed W's rather than W's that assume round genets

For control plots, we used a version of the model with all removal treatment effects set to zero. For removal treatment plots, we generated predictions from two versions of the model, with and without treatment effects set to zero. Because we were interested in comparing the predictions of these two models to observations, and were not interested in the effects of demographic stochasticity, we used a deterministic version of the model (e.g. recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a Poisson distribution with a mean of $\lambda$). We calculated the mean across quadrats of observed log cover change, $log(Cover_{t+1}/Cover_t)$ and compared it to the mean across quadrats of the predicted cover change.

@@

To quantify the long-term, population-level consequences of the removal effects, we built and then simulated an integral projection model (IPM). In our IPM, the population of species $j$ is represented by a density function $n(u_{j,t})$ which gives the density of genets of size $u$ at time $t$, with genet size on a natural-log scale, i.e. $n(u_{j,t})du$ is the number of genets whose area (on arithmetic scale) is between $exp(u_j)$ and $exp(u_j+du)$. The density function for size $v$ at time $t+1$ is given by
\begin{equation}
n(v_{j,t+1})=\int_{L_j}^{U_j} k_j (v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j))n(u_{j,t})   
\label{eqn:IPM}
\end{equation}
where the kernel $k_j (v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j))$ describes all possible transitions from size $u$ to $v$ and $\boldsymbol{\bar{W}}_j$  is a vector whose elements are the average crowding experienced from each of the other species by an individual of size $u_j$ of species $j$ (we describe below how ($\boldsymbol{\bar{W}}_j$ ) is calculated). The integral is evaluated over all possible sizes from a lower size limit $L$ to an upper size limit $U$ that extends past the range of observed sizes.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%%% Just for commenting
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[usenames]{color}
\newcommand{\new}{\textcolor{red}}
\newcommand{\comment}{\textcolor{blue}}

@@

\begin{enumerate}
\item A previous study of four dominant species from a sagebrush steppe community used historical data sets and multispecies population models to demonstrate weak interspecific interactions, implying very little competitive release. 
\item We conducted a removal experiment to test the model's ability to predict competitive release. We established new quadrats in the same location where the historical data were collected. We assigned half the new quadrats to a  perennial grass removal treatment and half to a treatment in which we removed the dominant shrub, \textit{Artemisia tripartita}.
\item We modeled survival, growth and recruitment as functions of local neighborhood species composition along with an indicator variable for removal treatment. For survival and recruitment, the removal treatment effects were never significantly positive, indicating that competitive release was not underestimated by the interspecific effects included in the ``baseline'' model. 
\item For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated competitive release. However, including information about the location of these grass individuals with respect to removed \textit{Artemisia} failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal.
\item For three of the four species, individual based models and integral projections models  based on the vital rate regressions showed that removal treatment effects had little impact on population growth rates or simulated equilibrium cover following a species removal. For \textit{P. spicata}, the population models that included effects of \textit{Artemisia} removal projected much stronger competitive release than the baseline models. 
\item \emph{Synthesis}: Although models based on observational data did not perfectly predict species responses to removal, overall the results increase our confidence that competitive release is in fact small for at least three of our four study species. The implication is that \new{interspecific} resource competition is not the primary factor determining the dynamics of this community.

@@

Stable coexistence requires that intraspecific interactions must limit populations more than interspecific interactions (Chesson 2000). 
\comment{(SPE: snip here)}
%In other words, when species are limited by different factors, intraspecific limitations will be stronger than interspecific limitations. 
%For example, species might differ in their use of multiple limiting resources (Tilman), respond differently to spatial and temporal variation in the
%environment (Chesson 2000), or host different natural enemies (Janzen, Connell). 
Whenever we study populations of species that appear to be coexisting stably, such as populations that have co-occurred in close proximity for extended periods, we should expect intraspecific limitations to be stronger than interspecific limitations. However, we still might expect interspecific interactions to be strong and negative.

@@

\new{Here we present a removal experiment, conducted in a sagebrush steppe} 
% We conducted a removal experiment in a sagebrush steppe 
to test whether our models, which are based on observational data, underestimate the strength of interspecific competition and competitive release. We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses. We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs. removal treatments after accounting for differences in local species composition? Second, how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data?

@@

The U.S. Sheep Experiment Station (USSES) is located 9.6 km north of Dubois, Idaho (44.2\degree N, 112.1\degree W), 1500 m above sea level. During the period of data collection (1926 – 1957), mean annual precipitation was 270 mm and mean temperatures ranged from -8\degree C (January) to 21\degree C (July). The vegetation is dominated by the shrub, \textit{Artemisia tripartita}, and the C3  perennial bunchgrasses \textit{Pseudoroegneria spicata}, \textit{Hesperostipa comata}, and \textit{Poa secunda}. These four species, the focus of our models, accounted for over 70\% of basal cover (grasses) and 60\% of canopy cover (shrubs and forbs).

Scientists at the USSES established 26 1-m$^2$ quadrats between 1926 and 1932. Eighteen quadrats were distributed among four ungrazed exclosures, and eight were distributed in two paddocks grazed at medium intensity spring through fall. All quadrats were located on similar topography and soils. In most years until 1957, all individual plants in each quadrat were mapped using a pantograph (Blaisdell 1958). The historical data set is public and available online \citep*{zachmann-etal-2010}. In 2007, we located 14 of the original quadrats, all of which are inside permanent livestock exclosures, and resumed annual mapped censusing using the traditional pantograph method.

We extracted data on survival, growth, and recruitment from the mapped quadrats based on plants' spatial locations. Our approach tracks genets, which may be composed of multiple polygons, as they fragment and/or coalesce. Each mapped polygon is classified as a surviving genet or a new recruit based on its spatial location relative to genets present in previous years (Lauenroth and Adler 2008). We modeled vital rates using data from 22 year-to-year transitions between 1926 and 1957, and eight year-to-year transitions from 2007 to 2015. Only four quadrates were observed for the first two transitions in the 1920's, while at least 14 quadrats were observed for all subsequent transitions.

\subsection*{Removal experiment}
In spring 2011, in a large exclosure containing six of the historical permanent quadrats, we selected locations for an additional 16 quadrats for the removal treatments. We used the following criteria to select the quadrat locations. Each plot had to have xx? cumulative cover/density? of the three target perennial grass species as well as yy? cover/density of \textit{Artemisia}. We mapped the quadrats in June, 2011 and then implemented the removal treatments in September (?), 2011. We randomly assigned 8 of the sixteen new quadrats to the sagebrush removal treatment, and the remaining 8 to the grass removal treatment. We cut sagebrush stems at ground level and applied herbicide to the cut stems. In the grass removal plots, we herbicided all perennial grasses. We removed plants zz cm beyond the quadrat boundaries to minimize edge effects. The sagebrush removal was virtually 100\% effective; grass removals required additional herbicide applications in April and May of each year, with fewer grasses remaining each additional year.

@@

Although the primary focus of our analysis is comparison of observed removal responses with responses predicted by our population models, we also conducted a more traditional statistical analysis of differences in cover trends on control vs. removal plots. The response variable for this analysis was year-to-year change in cover at the quadrat scale, calculated as $log(Cover_{t+1}/Cover_t)$. We tested for the effect of removal treatment on this measure of cover change with a linear mixed effects model implemented in the \textit{lme4} package (Bates et al. 2015) of \textit{R}, including year and quadrat as random effects. We considered the removal treatment effect to be significant when the 95\% confidence interval on the estimated coefficient did not overlap zero.

@@

Our regression models for survival, growth and recruitment follow previous work (Adler et al. 2010, Chu and Adler 2015) with a few exceptions, such as a new intercept term for the removal treatment. We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (`group'; here means a set of nearby quadrats located within one pasture or grazing exclosure), along with the removal treatment effect. Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatment $h$, from time $t$ to $t+1$  as
\begin{equation}
\mbox{logit}(S_{ijgh,t}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
\left \langle \boldsymbol{\omega}_{j,t}^S, \boldsymbol{W}_{ij,t} \right \rangle 
\label{eqn:survReg}
\end{equation}
where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial group,$\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival. $\boldsymbol{\omega}$ is a vector of interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$, by each species on the focal species. The vector $\boldsymbol{W}$ includes crowding from the four dominant species,  \textit{A. tripartita}, \textit{P. spicata}, \textit{H. comata}, and \textit{Poa secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points. 
\new{$\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
calculated as \texttt{sum(x*y)} in R.}

Our growth model has a similar structure. The change in genet size from time $t$ to $t+1$ , conditional on survival, is given by:
\begin{equation}
\mbox{logit}(u_{ijgh,t+1}) = \gamma_{j,t}^G + \varphi_{jg}^G+  \chi_{jh}^G  + \beta_{j,t}^G u_{ij,t} + 
\left \langle  \boldsymbol{\omega}_{j,t}^G, \boldsymbol{W}_{ij,t} \right \rangle + \varepsilon_{ij,t}^G .
\label{eqn:growReg}
\end{equation}
To capture non-constant error variance in growth, we modeled the variance $\varepsilon$  about the growth curve (\ref{eqn:growReg})  as a nonlinear function of predicted genet size:
\begin{equation}
Var(\varepsilon_{ij,t}^G) = a \exp ^{(bu_{ij,t+1})} .
\label{eqn:growVar}
\end{equation}

We model the crowding experienced by a focal genet as a function of the distance to and size of neighbor genets. In previous work, we assumed that the decay of crowding with neighbor distance followed a Gaussian function (Chu and Adler 2015), but here we use a data-driven approach (Teller et al. 2016). We model the crowding experienced by genet $i$ of species $j$ from neighbors of species $m$ as the sum of neighbor areas across a set of concentric annuli, $k$, centered at the plant,
\begin{equation}
w_{ijm,k} = F_{jm}(d_{k})A_{i,k}     
\label{eqn:wik}
\end{equation}
where $F_{jm}$ is the competition kernel (described below) for effects of species $j$ on species $m$, 
$d_{k}$ is the average of the inner and outer radii of annulus $k$, 
and $A_{im,k}$ is the total area of genets of species $m$ in annulus $k$ around genet $i$. The total crowding on 
genet $i$ exerted by species $m$ is
\begin{equation}
W_{ijm}  =\sum_k {w_{ijm,k}} .
\end{equation} 
Note that $W_{ijj}$ gives intraspecific crowding. The $W$'s are then the components of the $\boldsymbol{W}$ vectors 
introduced as covariates in the survival (\ref{eqn:survReg}) and growth (\ref{eqn:growReg}) regressions.

\new{We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, which means that distant plants have less effect 
than close plants. Otherwise, we let the data dictate the shape of the kernel by fitting a spline model 
using the methods of Teller et al. (2016). The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
and a smoothing parameter $\eta$ that controls the complexity of the fitted kernel. 
Demographic models such as \eqref{eqn:survReg} then have $\gamma$, $\varphi$, $\chi$ , 
$\beta$, $\boldsymbol{\omega}$, $\vec{b}$ and $\eta$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R}, 
by making the spline coefficients and $\eta$ the arguments of an objective function that computes $\boldsymbol{W}$ using the input coefficients, 
calls the model-fitting function \texttt{lmer} and/or \texttt{glmer} to fit the other parameters in the survival and growth regressions, and returns an approximate AIC value and model degrees of freedom ($df$) for survival and growth combined. We used the $\vec{b}$ values at the smoothest 
local minimum of AIC as a function of $df$, as in Teller et al. (2016). This approach assumes that one measure of crowding affects 
survival and growth. In addition, 
for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
within-species competition kernels $F_{jj}$. Based on previous work (Adler et al. 2010), we set all $F_{jm}$ equal to $F_{jj}$, meaning that 
the within-species competition kernel for species $j$ is also used to determine the effect of all other species on species $j$. We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels, which are shown in Fig. \ref{fig:CompKernels}. }
\comment{SPE: see Robin's comment about whether you mean $F_{mm}$ or $F_{jj}$; I'm not sure which one you did.}

Once we had estimated the competitions kernels, we used them to calculate the values of $\boldsymbol{W}$ for each individual, 
and fit the full survival and growth regressions, which include the interspecific interaction coefficients, $\boldsymbol{\omega}$. 
All genets in a quadrat were included in calculating $W$, but plants located within 5 cm of quadrat edges were not used in fitting. 
We fit the models using the \textit{R-INLA} package for \texttt{R} (Rue et al. 20xx).

We model recruitment at the quadrat level rather than at the individual genet level because the mapped data do not allow us to determine which recruits were produced by which potential parent genets. We assume that the number of individuals, $y$, of species $j$ recruiting at time $t+1$ in the location $q$ follows a negative binomial distribution:
\begin{equation}
y_{jq,t+1}= NegBin(\lambda_{jq,t+1},\theta) 	   
\label{eqn:recrDataModel}
\end{equation}
where $\lambda$ is the mean intensity and $\theta$ is the size parameter. In turn, $\lambda$ depends on the composition of the quadrat in the previous year:
\begin{equation}
\lambda_{jq,t+1} = C'_{jq,t} \exp{ \left(\gamma_{j,t}^R +  \varphi_{jg}^R + \chi_{jm}^R + 
\left \langle \boldsymbol{\omega}^R , \boldsymbol{\sqrt{C'}_{q,t}} \right \rangle \right) }
\label{eqn:recrProcessModel}
\end{equation}
where the superscript $R$ refers to Recruitment, $C'_{jq,t}$ is the `effective cover' (cm$^2$) of species $j$ in quadrat $q$ at time $t$, $\gamma$ is a time-dependent intercept, $\varphi$ is a coefficient for the effect of group location,
$\chi$ is the removal treatment effect, $\boldsymbol{\omega}$ is a vector of coefficients that determine the strength of intra- and interspecific density-dependence, and $\boldsymbol{C'}$ is the vector of ‘effective’ cover of each species in the community. Eqn (\ref{eqn:recrProcessModel}) is essentially the Ricker model for discrete time population growth (Ricker 1954), except that we use the square root of cover in the density-dependent term. We previously used a Ricker equation (Adler et al. 2010), but using the square-root transformation gave a lower Deviance Information Criterion (DIC) value (Spiegelhalter et al. 2002) for the recruitment model. Following previous work (Adler et al. 2010), we treated year and group variables as random factors, allowing intercepts to vary among years and spatial locations.

Due to the possibility that plants outside the mapped quadrat could contribute recruits to the focal quadrat or interact with plants in the focal quadrat, we estimated effective cover as a mixture of the observed cover, $C$, in the focal quadrat, $q$, and the mean cover, $\bar{C}$, across the group, $g$, in which the quadrat is located:
$C'_{jq,t}=p_j C_{jq,t}+(1-p_j) \bar{C}_{jg,t}$								
where $p$ is a mixing fraction between 0 and 1 that was estimated as part of fitting the model.

We ran Markov Chain Monte Carlo (MCMC) simulations in WinBUGS 1.4 (Lunn et al. 2000) to estimate the recruitment model parameters. Each model was run for 20,000 iterations and two independent chains with different initial values for parameters. We discarded the initial 10,000 samples. Convergence was observed graphically for all parameters, and confirmed by the Brooks-Gelman statistic (Brooks and Gelman 1998).

For all three vital rate models, we considered the effect of removal treatments to be significant when the 95\% confidence interval (or posterior
interval) on the estimate of $\chi$ did not overlap zero. In an effort to further understand the removal treatment effects, we explored additional variations on the growth model with a linear random effects model. For hypothesis testing with these models, we focused on the 95\% confidence interval of the coefficient of interest. We also compared the species composition of the control and removal plots at both the quadrat and neighborhood ($\boldsymbol{W}$) scales.

@@

The vital rate regressions provide a qualitative test of removal treatment effects: Does removal have an effect on a vital rate that goes above and beyond the effects of species interactions that are explicit in the model? However, the population-level response by each species integrates the effects of the removal treatment on all three vital rates. We used an individual-based model (IBM) to compare observed and predicted changes 
in population size (cover) from one year to the next.

To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the actual configuration of the four focal species (genet sizes and locations) observed in year $t$. We then projected the model forward one year, applying the random year effects corresponding to year $t$ and the spatial random effects corresponding to the quadrat of interest. Because we were interested in comparing model predictions to observations, and were not interested in the effects of demographic stochasticity, we used a deterministic version of the model (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a Poisson distribution with a mean of $\lambda$).
We first applied the recruitment regression to determine the expected number of new recruits at time $t+1$, which we then multiplied by the mean size of recruits observed for the focal species to get expected cover of new recruits. We then used the survival regression to predict the survival probability of each genet, and the growth regression to predict the change in size of each genet. Expected cover at $t+1$ is expected area of new recruits, plus the sum of the survival probabilities multiplied by the expected sizes. 
% need to REDO with observed W's rather than W's that assume round genets

For control plots, we used a version of the model with all removal treatment effects set to zero. For removal treatment plots, we generated predictions from two versions of the model, with and without treatment effects set to zero.  We calculated the mean across quadrats of observed log cover change, $log(Cover_{t+1}/Cover_t)$ and compared it to the mean across quadrats of the predicted cover change.

@@

To quantify the long-term, population-level consequences of the removal effects, we built and then simulated an integral projection model (IPM). In our IPM, the population of species $j$ is represented by a density function $n(u_{j,t})$ which gives the density of genets of size $u$ at time $t$, with genet size on a natural-log scale, i.e. $n(u_{j,t})du$ is the number of genets whose area (on arithmetic scale) is between $exp(u_j)$ and $exp(u_j+du)$. The density function for size $v$ at time $t+1$ is given by
\begin{equation}
n(v_{j,t+1})=\int_{L_j}^{U_j} k_j (v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j))n(u_{j,t})   
\label{eqn:IPM}
\end{equation}
where the kernel $k_j (v_j,u_j,\boldsymbol{\bar{W}}_j  (u_j))$ describes all possible transitions from size $u$ to $v$ and $\boldsymbol{\bar{W}}_j$  is a vector whose elements are the average crowding experienced from each of the other species by an individual of size $u_j$ of species $j$ (we describe below how $\boldsymbol{\bar{W}}_j$ is calculated). The integral is evaluated over all possible sizes from a lower size limit $L$ to an upper size limit $U$ that extends past the range of observed sizes.
